PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-03-31 manuscript_00.Rmd
alter data load

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)

@@

library(pool)
library(RSQLite)
meta_filter_v0 <- fst::read_fst("~/data/scEiaD/paper_data/2021_03_17_meta_filter.fst") %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                             TRUE ~ CellType))
# remove eric bharti 2D/3D RPE
meta_filter <- fst::read_fst("~/data/scEiaD_2022_02/2022_03_22_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                             TRUE ~ CellType),  
         Tissue = case_when(grepl('Iris', Tissue) ~ 'Iris',
                            TRUE ~ Tissue))

@@

## Updates to the scEiaD backend
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.90 web app in six important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300, which reduces the number of cells that pass QC but improves the quality. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.

@@

There are two interwoven reasons why PLAE is responsive, despite the huge amount of data it contains. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage. The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-21 sqlite database is abour 368 GB. Second, the visualizations in PLAE (e.g. the UMAP view) pull from the custom database directly and use highly optimized plotting packages (scattermore) to reduce drawing time. Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23). In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies.

@@

PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values. So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells. The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells. Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area.

@@

While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina. The structure of the cartoon retina was based on the human retina. This unique visualization displays the major cell types of the retina and the RPE / choroid layer behind it. The cell types are colored by the relative amounts of expression in the user selected gene. Like all visualizations in PLAE, the user can filter to only display data from certain metadata characteristics, like species or publication.

## Differential testing identifies thousands genes frequently enriched across various ocular cell types 
```{r}
scEiaD_2020_v01 %>% tbl('wilcox_diff_testing') %>% filter(Group == 'CellType (Predict)', FDR < 1e-5, mean_auc > 0.7 ) %>% group_by(Base) %>% summarise(Count = n()) %>% data.frame()
```

## Pan-study fovea vs peripheral photoreceptor identifies N genes consistently differentially expressed

>>>>

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
data_dir_vGiga <- '~/data/scEiaD/'
data_dir_vPLAE <- '~/data/scEiaD_2022_02/'

@@

library(pool)
library(RSQLite)
meta_filter_v0 <- fst::read_fst(glue("{data_dir_vGiga}/paper_data/2021_03_17_meta_filter.fst")) %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                             TRUE ~ CellType))

meta_filter <- fst::read_fst(glue("{data_dir_vPLAE}/meta_filter.fst")) %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                             TRUE ~ CellType),  
         Tissue = case_when(grepl('Iris', Tissue) ~ 'Iris',
                            TRUE ~ Tissue),

         Citation = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'Swamy VS, Fufa TD, Hufnagel RB, McGaughey DM Building the Mega Single Cell Transcriptome Ocular Meta-Atlas Gigascience 2021 Oct 13;10(10)',
                              TRUE ~ Citation),
         PMID = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ '34651173',
                          TRUE ~ PMID),
         study_accession = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'SRP329495',
                                     TRUE ~ study_accession))

@@

## Updates to the scEiaD backend
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.90 web app in six important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 200 to 300, which reduces the number of cells that pass QC but improves the quality. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.

@@

There are two interwoven reasons that PLAE is responsive, despite the huge amount of data present. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database allows for huge amounts of data to be stored for quick retrieval with minimal loading time. The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous. Second, the visualizations in PLAE (e.g. the UMAP view) pull from the custom database directly and use highly optimized plotting packages (scattermore) to reduce drawing time. Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study. In the same amount of time, PLAE can plot the cell type labels for over a million cells.

@@

PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values. So thus a user can, on the left, see Crx expression across over a million cells, and on the right, see cell type labels for those same million cells. The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells. Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area.

@@

While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.

## Differential testing identifies many genes consistently enriched in various ocular cell types 
```{r}
scEiaD_2020_v01 %>% tbl('wilcox_diff_testing') %>% filter(Group == 'CellType (Predict)', FDR < 1e-5, mean_auc > 0.7 ) %>% group_by(Base) %>% summarise(Count = n()) %>% data.frame()
```

## Fovea vs peripheral photoreceptor multi-study validation
